---
title: "htapp_methods_NSCLC_annotate.Rmd"
author: "Orr Ashenberg"
date: "4/21/2019"
output: html_notebook
---

Annotate NSCLC cancer data.

```{r}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())
library(Seurat)
library(dplyr)
library(Matrix)
library(gplots)
library(GSA)
library(stringr)
library(fossil)  # adjusted rand index function

# Set directories 
proj.path <- "/path/to/project/directory"
user.path <- "/path/to/user/files"

# Set date year_month_day to use in figure directory name and in names for saving Rda objects.
date <- "ENTER_DATE"  
name.project <- "HTAPP_NSCLC_example"

# Load Seurat object that was converted from cumulus.
load(pasteo(proj.path, "/src/Rdata/ENTER_DATE_HTAPP_NSCLC_example_convert_scrtools_to_seurat_cumulus.Rda")) 

# Read in cumulus results 
sampleid.cumulus <- c("NSCLC14_C4", "NSCLC14_PDEC", "NSCLC14_LE", "NSCLC14_combined")

# Create output directories.
figures.dir <- paste0(proj.path, "/results/NSCLC14/", date, "_10x_annotate/")
figures.path <- sapply(sampleid.cumulus, function(i) paste0(figures.dir, i, "_"), USE.NAMES = F) 

# Save cluster annotation scores.
Rda.path <- paste0(proj.path, "/src/Rdata/ENTER_DATE_NSCLC_exmample_seurat_annotate.Rda")
cluster.scores.prefix <- paste0(proj.path, "/src/Rdata/clusterscores/", date, "_10x_", name.project, "_")
cluster.scores.path <- paste(cluster.scores.prefix, sampleid.cumulus, "_cluster_scores.Rds", sep = "") 

out.dir <- c(figures.dir, paste0(proj.path, "/src/Rdata"), paste0(proj.path, "/src/Rdata/clusterscores"), paste0(figures.dir, "tSNE_scores/"))
sapply(out.dir, function(i) {if (!dir.exists(i)) {dir.create(i, recursive = T)} })


# Read in list of gene sets associated with different cell types.
file.gmt.S3A <- paste0(user.path, "/data/genelists/2018_LivnatJerby_melanoma_cell_types_Cell_S3A.gmt")
file.gmt.S3B <- paste0(user.path, "/data/genelists/2018_LivnatJerby_melanoma_cell_types_Cell_S3B.gmt")
file.gmt.lung <- paste0(user.path, "/data/genelists/Lung_Alex.gmt")
invisible(capture.output(gs.S3A <- GSA.read.gmt(file.gmt.S3A)))
invisible(capture.output(gs.S3B <- GSA.read.gmt(file.gmt.S3B)))
invisible(capture.output(gs.lung <- GSA.read.gmt(file.gmt.lung)))
gs.S3B$genesets <- gs.S3B$genesets[-5]  # remove malignant cell signature
gs.S3B$geneset.names <- gs.S3B$geneset.names[-5]
gs.S3B$geneset.descriptions <- gs.S3B$geneset.descriptions[-5]
gs.S3B$genesets <- c(gs.S3B$genesets, gs.lung$genesets[1:9])
gs.S3B$geneset.names <- c(gs.S3B$geneset.names, gs.lung$geneset.names[1:9])
gs.S3B$geneset.descriptions <- c(gs.S3B$geneset.descriptions, gs.lung$geneset.descriptions[1:9])

# genes <- gs.S3B$genesets[gs.S3B$geneset.names == "CAF"][[1]]
# genes <- c("TPSAB1", "TPSB2", "CPA3", "KIT", "FCER1G") # Mast cell markers
# genes <- c("CD3D", "TRAC", "CD4", "CD8A", "NCAM1") # T cell markers
# genes <- c("CD19", "MS4A1", "CD79A", "JCHAIN", "MZB1", "IGHM") # B cell markers

# Load my functions.
source(paste0(user.path, "/code/R/plotutils.R"))
source(paste0(user.path, "/code/R/seuratutils.R"))
source(paste0(user.path, "/code/R/color.R"))
```

## Get Livnat Jerby marker genes and make gmt files
```{r, eval = F}
library(readxl)

file.marker <- paste0(proj.path, "/doc/genesets/2018_LivnatJerby_melanoma_cell_types_Cell.xlsx")

excel <- read_excel(file.marker, sheet = "TableS3A_cell.type.markers", skip = 1)
lines <- c()
for (col in colnames(excel)) {
  genes <- excel[[col]][!is.na(excel[[col]])]
  line <- paste0(col, "\t", col, "\t", paste0(genes, collapse = "\t"))
  lines <- c(lines, line)
}
fileConn <- file(paste0(proj.path, "/doc/genesets/2018_LivnatJerby_melanoma_cell_types_Cell_S3A.gmt"))
writeLines(lines, fileConn)
close(fileConn)

excel <- read_excel(file.marker, sheet = "TableS3B_denovo.cell.type.sig", skip = 1)
lines <- c()
for (col in colnames(excel)) {
  genes <- excel[[col]][!is.na(excel[[col]])]
  line <- paste0(col, "\t", col, "\t", paste0(genes, collapse = "\t"))
  lines <- c(lines, line)
}
fileConn <- file(paste0(proj.path, "/doc/genesets/2018_LivnatJerby_melanoma_cell_types_Cell_S3B.gmt"))
writeLines(lines, fileConn)
close(fileConn)
```

## Clustering 
Seurat uses graph-based clusering with K-nearest neighbor graph and Euclidean distances in PCA space. The resolution parameter can be altered between 0.6-1.2 and increasing it increases the number of clusters.

```{r cluster, eval = T}
gcdata <- sapply(gcdata, function(data) FindClusters(object = data, reduction.type = "pca", dims.use = 1:20, resolution = 1, print.output = F, save.SNN = T, force.recalc = T))
PrintFindClustersParams(object = gcdata[[1]])

# Find cluster biomarkers and store in list of dataframes
gcdata.markers <- lapply(gcdata, function(data) FindAllMarkers(object = data, min.pct = 0.2, logfc.threshold = 0.5, max.cells.per.ident = 1500, test.use = test, return.thresh = 0.1/nrow(data@data), print.bar = T))
top2genes <- lapply(gcdata.markers, function(data) data %>% group_by(cluster) %>% top_n(2, avg_logFC))
top10genes <- lapply(gcdata.markers, function(data) data %>% filter(avg_logFC > 0) %>% group_by(cluster) %>% top_n(10, avg_logFC))
top100genes <- lapply(gcdata.markers, function(data) data %>% group_by(cluster) %>% top_n(100, avg_logFC))

# Write cluster markers to csv file
sapply(seq_along(gcdata), function(i) write.csv(top100genes[[i]], file=paste0(figures.path[i], "top_DE_genes.csv")))
```

## Annotate cell clusters
We want to determine what cell type each cell cluster corresponds to. We will read in sets of genes known to be associated with a certain cell type or cell state and score each cell by that set of genes. We can then take the average score for that gene set for all cells within the cluster.

```{r annotate, eval = F}
# We want to determine what cell type each cell cluster corresponds to. We will read in sets of genes known to be associated with a certain cell type or cell state and score each cell by that set of genes. We can then take the average score for that gene set for all cells within the cluster.
name.cluster <- "louvain_labels"
gs <- gs.S3B

# Iterate over each Seurat object and score each cluster with gene signatures.
for (s in seq(sampleid.cumulus)) {
  out.prefix <- paste0(figures.dir, "tSNE_scores/", sampleid.cumulus[s], "/")
  cluster.scores <- RankModuleScore(gcdata[[s]], name.cluster, gs, out.prefix)
  # For each cluster, rank the gene modules by their mean score for that cluster.
  for (i in seq(cluster.scores[, name.cluster])) {
    # Order gene modules in descending order, and ignore the first column which is the cluster number
    top.clusters <- names(cluster.scores)[order(-cluster.scores[i, -c(1)]) + 1]
    test.df <- cbind(data.frame(cluster = cluster.scores[i, name.cluster]), cluster.scores[i, top.clusters[1:3]])
    print(test.df)
  }
  saveRDS(cluster.scores, cluster.scores.path[s])
}

# Example of reading in saved cluster scores and printing them.
i <- 1
name.cluster <- "louvain_labels"
cluster.scores <- readRDS(cluster.scores.path[i])
for (i in seq(cluster.scores[, name.cluster])) {
  # Order gene modules in descending order, and ignore the first column which is the cluster number
  top.clusters <- names(cluster.scores)[order(-cluster.scores[i, -c(1)]) + 1]
  test.df <- cbind(data.frame(cluster = cluster.scores[i, name.cluster]), cluster.scores[i, top.clusters[1:3]])
  print(test.df)
}
```
## Label cell types with annotation
Label cell subsets based on scoring gene modules above.
```{r cluster_analysis, eval = F}


PlotAnnotation <- function(gcdata, file.plot, title) {
  df <- data.frame(gcdata@dr$umap@cell.embeddings, ident = gcdata@ident)
  n.clusters <- n_distinct(df$ident)
  centers <- df %>% group_by(ident) %>% summarise(x = median(UMAP1), y = median(UMAP2))  # cluster labels
  p <- ggplot(data = df) + geom_point(mapping = aes(UMAP1, UMAP2, color = ident), size = 1, show.legend = F) +
    labs(color = "cluster") + scale_colour_manual(values = material.heat(n.clusters)) +
    geom_label(data = centers, mapping = aes(x = x, y = y, label = ident), size = 5,
              color = "black", show.legend = F, alpha = 0.75) +
    guides(colour = guide_legend(override.aes = list(size=5))) + ggtitle(title)
  print(p)
  ggsave(file.plot, width = 8, height = 8)
}

# Identify cell types based on marker genes and annotate clusters by their cell type
i <- 1
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:12
new.cluster.ids <-c("Macrophage", "Epithelial cell", "Macrophage", "Macrophage", "Macrophage", "T cell", "Macrophage", "Epithelial cell", "Epithelial cell", "Macrophage", "Epithelial cell", "B cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

i <- 2  # Further divide fibroblasts and verify mast cells
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:13
new.cluster.ids <-c("Endothelial cell", "Epithelial cell", "Fibroblast", "T cell", "Fibroblast", "Macrophage", "Macrophage", "Macrophage", "Fibroblast", "T cell", "B cell", "Fibroblast", "Mast cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

i <- 3
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:14
new.cluster.ids <-c("Endothelial cell", "T cell", "Fibroblast", "Fibroblast", "Macrophage", "Epithelial cell", "Fibroblast", "Macrophage", "Macrophage", "Fibroblast", "Endothelial cell", "Endothelial cell", "B cell", "Mast cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

i <- 4
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:20
new.cluster.ids <-c("Endothelial cell", "T cell", "Macrophage", "Macrophage", "Epithelial cell", "Fibroblast", "Fibroblast", "Macrophage", "Epithelial cell", "Fibroblast", "Macrophage", "T cell", "Macrophage", "Macrophage", "Mast cell", "B cell", "Endothelial cell", "Endothelial cell", "Epithelial cell", "B cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

i <- 5
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:10
new.cluster.ids <-c("T cell", "T cell", "T cell", "Macrophage", "B cell", "B cell", "Epithelial cell", "Fibroblast", "Mast cell", "Endothelial cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

i <- 6
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:14
new.cluster.ids <-c("Epithelial cell", "T cell", "Macrophage", "T cell", "Macrophage", "Fibroblast", "B cell", "T cell", "Epithelial cell", "Endothelial cell", "B cell", "Macrophage", "Epithelial cell", "Mast cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

i <- 7
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:16
new.cluster.ids <-c("Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Fibroblast", "Epithelial cell", "Fibroblast", "Endothelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

i <- 8
file.plot <- paste0(figures.path[i], "tSNE_annotate.png")
title <- paste0(sampleid.cumulus[i])
current.cluster.ids <- 1:19
new.cluster.ids <-c("Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "T cell", "Epithelial cell", "Fibroblast", "Epithelial cell", "Macrophage", "Fibroblast", "Endothelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "Epithelial cell", "B cell", "B cell", "Mast cell")
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "louvain_labels")
gcdata[[i]]@ident <- plyr::mapvalues(x = gcdata[[i]]@ident, from = current.cluster.ids, to = new.cluster.ids)
gcdata[[i]] <- StashIdent(gcdata[[i]], save.name = "annotate")
PlotAnnotation(gcdata[[i]], file.plot, title)

# Save current progress with annotating each sample.
save(gcdata, sampleid.cumulus, file = Rda.path)
```

## Cluster analysis
We want to compare the differences in cell cluster composition across the different experimental conditions. For example, does one condition result in higher representation of immune cells? We make tables of experimental conditions vs number of cells in each cell cluster. We then look at this in a few ways.
1: Within an experimental condition, what percent of cells are T, B, NK, etc...?
2: Within a cell subset like T cells or B cells, what percent of those cells come from each experimental condition?
3: Given the background distribution of cell types across all the experimental conditions, within an experimental condition is a cell type over or under-represented? In other words, does one experimental condition do better or worse at detecting a specific cell type compared to the average across all experimental conditions? 
```{r cluster_freq, eval = T}
# Use annotated cell subset labels.
# gcdata@ident <- plyr::mapvalues(x = gcdata@ident, from = current.cluster.ids, to = new.cluster.ids)
# gcdata <- gcdata.liger
i <- 4
gcdata[[i]] <- SetAllIdent(gcdata[[i]], id = "annotate")

# Cluster vs Experimental condition where row is cluster and column is experimental condition.
table.cluster.experiment <- table(gcdata[[i]]@ident, gcdata[[i]]@meta.data$Condition)

# Within a cell cluster, what percent of cells come from each experimental condition?
# Within an experimental condition, what percent of cells are cell type 1,2,3,etc...?
freq.cluster.experiment <- table.cluster.experiment / rowSums(table.cluster.experiment) 
freq.experiment.cluster <- sweep(table.cluster.experiment, 2, colSums(table.cluster.experiment), '/') 

# Across all cells, the frequency of cells coming from each experiment.
# Across all cells, the frequency of cells coming from each cell cluster.
freq.experiment <- prop.table(colSums(table.cluster.experiment))
freq.cluster <- prop.table(rowSums(table.cluster.experiment))  

# Make bar plot of frequency of cell clusters found within each experiment.
df <- as.data.frame.matrix(freq.experiment.cluster)  # convert table to dataframe, and then gather key-values
df["cluster"] <- rownames(df)
df <- gather(df, "experiment", "frequency", c(1:(ncol(df)-1)))
df$experiment <- factor(df$experiment, levels = levels(gcdata[[i]]@meta.data$Condition))  # order samples
df$cluster <- factor(df$cluster, levels = unique(df$cluster))  # order by original cluster ordering
ggplot(data = df) + geom_bar(mapping = aes(x = experiment, y = frequency, fill = cluster), stat = "identity") + theme(axis.text.x  = element_text(angle=90, hjust=1)) + scale_fill_manual(values=material.heat(length(levels(df$cluster))))
ggsave(paste0(figures.path[i], 'cluster_freq_bar_plot.png'), width = 6, height = 8)

# Make bar plot of frequency of experimental conditions found within each cell cluster.
df <- as.data.frame.matrix(freq.cluster.experiment)  # convert table to dataframe, and then gather key-values
df["cluster"] <- rownames(df)
df <- gather(df, "experiment", "frequency", c(1:(ncol(df)-1)))
df$experiment <- factor(df$experiment, levels = levels(gcdata[[i]]@meta.data$Condition))  # order samples
df$cluster <- factor(df$cluster, levels = unique(df$cluster))  # order by original cluster ordering
ggplot(data = df) + geom_bar(mapping = aes(x = cluster, y = frequency, fill = experiment), stat = "identity") + theme(axis.text.x  = element_text(angle=90, hjust=1)) + scale_fill_manual(values=material.heat(length(levels(df$experiment))))
ggsave(paste0(figures.path[i], 'experiment_freq_bar_plot.png'), width = 6, height = 8)
```
